---
output: md_document
editor_options: 
  chunk_output_type: console
---

When working with distributions in R, each distribution has four functions, namely:

* `dXXX` - density function.
* `rXXX` - generate random number from this distribution.
* `pXXX` - returns the area to the left of the given value.
* `qXXX` - returns the quantile for the given value/area.

Where `XXX` is the distribution name (e.g. `norm`, `binom`, `t`, etc.).

The `VisualStats::plot_distributions()` function will generate four plots representing the four R 
distribution functions. For each subplot points correspond to the first parameter of the 
corresponding function (note the subplot for the random `rXXX` function does not have points
since this simply returns random values from that distribution). The arrows correspond to 
what that function will return.

The top two plots (`dXXX` and `rXXX`) plot the distribution. The bottom two plots are the
cumulative density function for the given distribution. The CDF describes the probability that a 
random variable (X) will be less than or equal to a specific value (`x`), written as 
F(x) = P(X ≤ x). The CDF provides a complete view of a random variable's distribution by 
accumulating probabilities up to that point.

### Available distributions

```{r, echo=FALSE, results='asis'}
data("distributions", package = 'VisualStats')
for(i in 1:nrow(distributions)) {
	cat(paste0('* [', distributions[i,]$label, '](', distributions[i,]$Wikipedia, '): ', 
			   '`X', distributions[i,]$function_name, '`',
			   '\n\n'))
}
```

```{r, eval=FALSE}
helptext <- help(dt, package = stats)
tools:::Rd2txt(utils:::.getHelpFile(as.character(helptext)))
db <- tools::Rd_db("stats")
tmp <- lapply(db, tools:::.Rd_get_metadata, "description")
tmp <- lapply(db, tools:::.Rd_get_metadata, "usage")
tmp$Normal.Rd

```

