---
title: "VisualStats: R package for visualizing statistical tests"
output: github_document
editor_options:
  chunk_output_type: console
---

Website: https://jbryer.github.io/VisualStats/

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE, error=FALSE}
library(devtools)
document()
install(build_vignettes = TRUE, dependencies = FALSE)
```

This package contains functions and [Shiny](https://shiny.rstudio.com) applications designed to visualize statistical tests. The latest version can be installed from Github using the `remotes` package:

```{r, eval=FALSE}
remotes::install_github('jbryer/VisualStats')
```

The vignettes currently available:

```{r}
library(VisualStats)
vignette(package = 'VisualStats')[['results']][,c('Item', 'Title'), drop = FALSE]
```

Shiny apps available. You can run the apps using the `shiny_demo('APP_NAME')` where `APP_NAME` is from the list below.

```{r}
shiny_demo()
```

All available functions and datasets:

```{r}
ls('package:VisualStats')
```
